---
title: "Cornwall Fire Model"
author: "James Millington"
format: gfm
editor: visual
---

```{r}
#| warning: false
library(terra)
library(tidyverse)
library(googlesheets4)
```

## Introduction

This document presents code and to implement the McArthur Fire model for a region of Cornwall, UK, parameterised using field data collected by King's College London students in the Dept. of Geography.

## Model Parameterisation

Equations for the McArthur fire model we will use come from [Noble *et al.* (1980)](https://doi.org/10.1111/j.1442-9993.1980.tb01243.x).

In the field, students collected data to help estimate key variables for McArthur's model for different land cover types represented in the UK CEH Land Cover Map:

-   degree of curing, *C*, in %

-   fuel load, *W*, in tons per hectare

By combining these variables with climate variables and the UK CEH Land Cover Map we can estimate fire danger and rate of spread across Cornwall and for different climate scenarios.

Summaries of the data collected by the six groups of students are shown in the Summary tab of [the Google spreadsheet used to record the data](https://docs.google.com/spreadsheets/d/1HReAFVljlUpjqmHOGYqvDZcUn9lkkPPRQ9ebb5BkQso/edit?usp=sharing). In turn, these summaries depend on summary equations in each of the data sheets (see bottom of data). Finally, calculations from the combined student field measurements are made in cells A26:E30 (shaded blue) on the Summary tab. Several of these calculations are based on the information presented in [Pasalodos-Tato *et al.* (2015)](http://doi.org/10.1007/s10342-015-0870-6).

We can read these calculated parameter values for *C* and *W* directly our model code here:

```{r}
#| warning: false
sheet_data <-range_read(ss="https://docs.google.com/spreadsheets/d/1HReAFVljlUpjqmHOGYqvDZcUn9lkkPPRQ9ebb5BkQso/edit#gid=937321199",
                        sheet="Summary",
                        range="A26:E30")
sheet_data
```

## Model Application

```{r}
fpath <- "/home/james/OneDrive/Teaching/2022-23/undergrad/fieldwork/cornwall/data/lcm-2020-25m_tiled_4639069/sx/sx06.tif"
```

## References

Pasalodos-Tato, M., Ruiz-Peinado, R., Del Río, M., & Montero, G. (2015). Shrub biomass accumulation and growth rate models to quantify carbon stocks and fluxes for the Mediterranean region. *European Journal of Forest Research*, *134*, 537-553. <https://doi.org/10.1007/s10342-015-0870-6>

Noble, I. R., Gill, A. M., & Bary, G. A. V. (1980). McArthur's fire‐danger meters expressed as equations. *Australian Journal of Ecology*, *5*(2), 201-203. <https://doi.org/10.1111/j.1442-9993.1980.tb01243.x>
