---
title: "Cornwall Fire Model"
author: "James Millington"
format: gfm
editor: visual
---

```{r}
#| warning: false
library(terra)
library(tidyverse)
library(googlesheets4)
```

## Introduction

This document presents code and to implement the McArthur Fire model for a region of Cornwall, UK, parameterised using field data collected by King's College London students in the Dept. of Geography.

## Model Inputs

Equations for the McArthur fire model we will use come from [Noble *et al.* (1980)](https://doi.org/10.1111/j.1442-9993.1980.tb01243.x). To apply the model to a particular location and time we need to provide various numeric inputs.

In the field, students collected data to help estimate key variables for McArthur's model for different land cover types represented in the UK CEH Land Cover Map:

-   degree of curing, *C*, in %

-   fuel load, *W*, in tons per hectare

By combining these variables with climate variables and the UK CEH Land Cover Map we can estimate fire danger and rate of spread across Cornwall and for different climate scenarios.

Summaries of the data collected by the six groups of students are shown in the Summary tab of [the Google spreadsheet used to record the data](https://docs.google.com/spreadsheets/d/1HReAFVljlUpjqmHOGYqvDZcUn9lkkPPRQ9ebb5BkQso/edit?usp=sharing). In turn, these summaries depend on summary equations in each of the data sheets (see rows at the bottom of input data). Finally, calculations from the combined student field measurements are made in cells A26:E30 (shaded blue) on the Summary tab. Several of these calculations are based on the information presented in [Pasalodos-Tato *et al.* (2015)](http://doi.org/10.1007/s10342-015-0870-6).

We can read these estimated values for *C* and *W* for the different land cover types directly our model code here:

```{r}
#| warning: false
sheet_data <-range_read(ss="https://docs.google.com/spreadsheets/d/1HReAFVljlUpjqmHOGYqvDZcUn9lkkPPRQ9ebb5BkQso/edit#gid=937321199",
                        sheet="Summary",
                        range="A26:E30")
sheet_data
```

## Model Application

```{r}
fpath <- "/home/james/OneDrive/Teaching/2022-23/undergrad/fieldwork/cornwall/data/lcm-2020-25m_tiled_4639069/sx/sx06.tif"
sx06 <- rast(fpath, lyrs=1)
plot(sx06)
```

The colours and their corresponding values in the map above correspond to the UK CEH Land Cover classification.

We shall re-classify these

| Fire Fuel Type | Fire Fuel Type ID | Original CEH Classes |
|----------------|-------------------|----------------------|
| Deciduous      | 1                 | 1                    |
| Conifer        | 2                 | 2                    |
| Arable         | 3                 | 3                    |
| Grassland      | 4                 | 4-7                  |
| Heathland      | 5                 | 9-10                 |
| Non-Veg        | 6                 | 8, 11-21             |

: Re-classification of land cover for fire model

```{r}
decid <- c(1,1,1)
conif <- c(2,2,2)
arable <- c(3,3,3)
grass <- c(4,7,4)
nonVeg <- c(8,8,6)
heath <- c(9,10,5)
nonVeg2 <- c(11,21,6)

FMC <- rbind(decid, conif, arable, grass, nonVeg, heath, nonVeg2)

fuelType <- classify(sx06, rcl=FMC, right=NA)

plot(fuelType, main="Fire Fuel Types")
```

Now we create raster maps for the curing and biomass inputs, applying the estimates from the field to each of our land cover types. First, the curing input (from % dead vegetation) for Grassland and Heathland.

```{r}
curing <- fuelType
curing[curing==1]<-NA
curing[curing==2]<-NA
curing[curing==3]<-NA
curing[curing==4]<-as.numeric(sheet_data[1,2])
curing[curing==5]<-as.numeric(sheet_data[1,3])
curing[curing==6]<-NA
plot(curing, main="Curing (% dead veg)")
```

Next biomass for Deciduous forest , Confier forest, Grassland and Heathland.

```{r}
biomass <- fuelType
biomass[biomass==1]<-as.numeric(sheet_data[4,4])
biomass[biomass==2]<-as.numeric(sheet_data[4,5])
biomass[biomass==3]<-NA
biomass[biomass==4]<-as.numeric(sheet_data[4,2])
biomass[biomass==5]<-as.numeric(sheet_data[4,3])
biomass[biomass==6]<-NA
plot(biomass, main="Biomass (tons per ha)")
```

We also need to provide climate inputs - if we assume climate is uniform across the study region we don't need to make rasters but can just create individual variables.

```{r}
temperature <- 20  #degrees C
humidity <- 50     #%
wind <- 10         #km/hr
KDBI <- 400     #KDBI (range is 0 [no drought] to 800 [extreme drought])
daysRain <- 1   #days since rain
pptn <- 10      #last rain in mm
  
drought <- 0.191*(KDBI+104)*(daysRain+1)^1.5/(3.52*(daysRain+1)^1.5+pptn-1)
```

Now we can calculate Fire Danger using the McArthur grassland fire danger model for Grassland and Heathland fuel types (land covers) using their respective curing and biomass values, and the forest fire danger model for the Deciduous and Conifer types (based solely on climate inputs).

```{r}
GrassM <- (((97.7+4.06*humidity)/(temperature+6)-0.00854*humidity))*((100-curing)/100)
GrassFDM <- 3.35*biomass*exp(-0.0897*GrassM+0.0403*wind)
ForestFDM <- 1.25*drought*exp((temperature-humidity)/30+0.0234*wind)


FDM <- GrassFDM
FDM[fuelType==1]<-ForestFDM    #deciduous
FDM[fuelType==2]<-ForestFDM    #conifer
FDM[fuelType==3]<-max(GrassFDM[fuelType==4])/2   #half grassland danger 
FDM[fuelType==6]<-0    #non-veg

plot(FDM, main="Fire Danger (McArthur scale 0-100")
```

We can also calculate predicted Rate of Spread (RoS) once a fire is alight. For forest fuel types this is dependent on a combination of fire danger and biomass, whereas for fuel types based on the grass model (Grassland, Heathland, Arable) RoS is related only to fire danger.

```{r}
RoS <- FDM

forestMulti<-max(FDM[fuelType==1])*0.0012

grassRoS<-max(FDM[fuelType==4]*0.13)
heathRoS<-max(FDM[fuelType==5]*0.13)

RoS[fuelType==1]<-biomass*forestMulti
RoS[fuelType==2]<-biomass*forestMulti
RoS[fuelType==3]<-grassRoS
RoS[fuelType==4]<-heathRoS
RoS[fuelType==5]<-grassRoS
RoS[fuelType==6]<-0

plot(RoS, main="Rate of Spread (km per hour)")
```

We could now use this model to explore different scenarios, primarily of different climatic conditions but also for different possible land use/cover futures.

To explore the effect of climate, set values we are interested in:

```{r}
temperature <- 30  #degrees C
humidity <- 40     #%
wind <- 10         #km/hr
KDBI <- 200     #KDBI (range is 0 [no drought] to 800 [extreme drought])
daysRain <- 10   #days since rain
pptn <- 10      #last rain in mm
```

Then calculate the fire danger for these new climate values:

```{r}
drought <- 0.191*(KDBI+104)*(daysRain+1)^1.5/(3.52*(daysRain+1)^1.5+pptn-1)
GrassM <- (((97.7+4.06*humidity)/(temperature+6)-0.00854*humidity))*((100-curing)/100)
GrassFDM <- 3.35*biomass*exp(-0.0897*GrassM+0.0403*wind)
ForestFDM <- 1.25*drought*exp((temperature-humidity)/30+0.0234*wind)
```

And finally apply the model to our map of fuel types to estimate fire danger

```{r}
FDM <- GrassFDM
FDM[fuelType==1]<-ForestFDM    #deciduous
FDM[fuelType==2]<-ForestFDM    #conifer
FDM[fuelType==3]<-max(GrassFDM[fuelType==4])/2   #half grassland danger 
FDM[fuelType==6]<-0    #non-veg

plot(FDM, main="McArthur Fire Danger")
```

And RoS for the new climate values

```{r}
RoS <- FDM

forestMulti<-max(FDM[fuelType==1])*0.0012

grassRoS<-max(FDM[fuelType==4]*0.13)
heathRoS<-max(FDM[fuelType==5]*0.13)

RoS[fuelType==1]<-biomass*forestMulti
RoS[fuelType==2]<-biomass*forestMulti
RoS[fuelType==3]<-grassRoS
RoS[fuelType==4]<-heathRoS
RoS[fuelType==5]<-grassRoS
RoS[fuelType==6]<-0

plot(RoS, main="Rate of Spread (km per hour)")
```

## References

Pasalodos-Tato, M., Ruiz-Peinado, R., Del Río, M., & Montero, G. (2015). Shrub biomass accumulation and growth rate models to quantify carbon stocks and fluxes for the Mediterranean region. *European Journal of Forest Research*, *134*, 537-553. <https://doi.org/10.1007/s10342-015-0870-6>

Noble, I. R., Gill, A. M., & Bary, G. A. V. (1980). McArthur's fire‐danger meters expressed as equations. *Australian Journal of Ecology*, *5*(2), 201-203. <https://doi.org/10.1111/j.1442-9993.1980.tb01243.x>
